---
title: "Week 3 Lab Exercise: R Fundamentals -- Data Types & Structures"
subtitle: "QMIR 2026"
author: ""
date: last-modified
date-format: MMMM D, YYYY
format:
  pdf:
    toc: true
    number-sections: true
    papersize: a4
    fontsize: 12pt
---

# Introduction

This document accompanies the Week 3 lecture. We will work through it together step by step. The goal is not to finish as fast as possible -- it is to build good habits: writing clean code, checking your output, and understanding what each line does before moving on.

At the bottom of the document you will find **optional exercises** if you finish early.

# Part I: Objects, Functions & Pipes

## Objects -- Storing Information

We start by creating a few objects. An object has a name, a value, and a type.

```{r}
# Assign some values to named objects
country <- "Germany"
gdp_per_capita <- 54300
is_eu_member <- TRUE
```

Now print each object by typing its name:

```{r}
country
gdp_per_capita
is_eu_member
```

**What to notice:** nothing is printed when you *assign* — only when you *call* the name.

## Functions -- Doing Things with Objects

Functions take inputs (arguments) and return outputs. We apply them to the objects we just created.

```{r}
gdp_values <- c(42300, 54300, 33200, 28700, 19800)

mean(gdp_values)
sqrt(gdp_per_capita)
nchar(country)
```

We can also write our own small functions:

```{r}
add_vat <- function(price) {
  price * 1.19
}

add_vat(100)
add_vat(gdp_per_capita)
```

## Getting help

```{r}
help(mean)
?mean()
```

## The Pipe -- Chaining Operations

Without the pipe, nested calls become hard to read:

```{r}
round(mean(sqrt(gdp_values)), digits = 2)
```

With the native pipe `|>`, we read left to right:

```{r}
gdp_values |> sqrt() |> mean() |> round(digits = 2)
```

**The pipe says:** *take this, and then do that.*

# Part II: Data Types

## Numeric

```{r}
gdp <- 54300.5 # double -- decimal numbers
seats <- 736L # integer -- whole numbers (note the L)

class(gdp)
class(seats)
```

## Character

```{r}
capital <- "Berlin"
country <- "Germany"

nchar(country)
toupper(country)
paste(country, "is a member of the EU")
```

Numbers stored as characters are **not** numbers -- R will not do math with them:

```{r}
#| error: true
"42" + 1
```

But we can convert them:

```{r}
as.numeric("42") + 1
```

## Logical

```{r}
is_eu_member <- TRUE
population <- 84000000

population > 50000000
population == 84000000
!is_eu_member
```

Try combining conditions:

```{r}
population > 50000000 & is_eu_member
population > 50000000 | is_eu_member
```

## Factors and Dates

Factors are for categorical variables with a fixed set of levels:

```{r}
regime <- factor(
  c("democracy", "autocracy", "democracy", "hybrid"),
  levels = c("autocracy", "hybrid", "democracy")
)

regime
levels(regime)
```

Dates allow calendar arithmetic:

```{r}
today <- Sys.Date()
treaty_of_rome <- as.Date("1957-03-25")

today - treaty_of_rome
```



# Part III: Data Structures

## Vectors

Vectors are the basic building block in R. All elements must be the same type.

```{r}
countries <- c("Germany", "France", "Italy", "Spain", "Poland")
population <- c(84, 68, 60, 47, 38) # millions
eu_member <- c(TRUE, TRUE, TRUE, TRUE, TRUE)

length(countries)
class(population)
```

What happens when we mix types?

```{r}
mixed <- c(1, 2, "three")
mixed
class(mixed)
```

**What to notice:** R silently converts everything to the most flexible type. This is called *coercion* -- it can cause quiet bugs if you are not paying attention.

## Data Frames

A data frame is the workhorse of data analysis in R. Each column is a vector; columns can be different types.

```{r}
eu_data <- data.frame(
  country = c("Germany", "France", "Italy", "Spain", "Poland"),
  population = c(84, 68, 60, 47, 38),
  eu_member = c(TRUE, TRUE, TRUE, TRUE, TRUE),
  regime = c("democracy", "democracy", "democracy", "democracy", "democracy")
)

eu_data
```

Explore the data frame:

```{r}
nrow(eu_data)
ncol(eu_data)
head(eu_data)
str(eu_data)
summary(eu_data)
```

## Lists

Lists are flexible containers -- they can hold anything, including objects of different types and lengths.

```{r}
country_info <- list(
  name = "Germany",
  population = 84000000,
  eu_member = TRUE,
  parties = c("SPD", "CDU", "Greens", "FDP")
)

country_info
```

You will encounter lists constantly as outputs of statistical models later in the course.



# Part IV: Accessing & Exploring Data

## Indexing Vectors with `[ ]`

```{r}
countries[1] # first element
countries[c(1, 3)] # first and third
countries[-2] # everything except the second
```

Logical indexing -- filter by condition:

```{r}
population > 50
countries[population > 50]
```

## Indexing Data Frames with `[ ]` and `$`

```{r}
eu_data[1, ] # first row, all columns
eu_data[, 2] # all rows, second column
eu_data[1, 2] # specific cell: row 1, column 2
```

Access a column by name with `$` -- this returns a vector:

```{r}
eu_data$country
eu_data$population
```

Combine: filter rows by a condition:

```{r}
eu_data[eu_data$population > 60, ]
```



# Optional Exercises

These exercises are for students who finish the walkthrough early. Work through them in order. Write a short comment or prose sentence explaining what each block of code does.



## Optional 1 -- Build Your Own Data Frame

Create a data frame about a topic of your choice (e.g., political parties, countries, historical events). It should have:

- At least 5 rows
- At least 4 columns, including at least one `character`, one `numeric`, and one `logical` column

Then run `str()` and `summary()` on it.

```{r}
# Your code here

```



## Optional 2 -- Factors and Ordering

Take a character vector of regime types (e.g., `"democracy"`, `"autocracy"`, `"hybrid"`) and convert it to a factor with a meaningful level order. Then:

- Print the factor
- Use `table()` to count how often each level appears
- Try `sort()` on the factor and observe what happens

```{r}
# Your code here

```



## Optional 3 -- Logical Filtering

Using the `eu_data` data frame from the walkthrough (or your own), write three different filtering expressions using `[ ]` and logical conditions. For example:

- Countries with population above a threshold
- Countries where `eu_member` is `TRUE`
- Combine two conditions with `&` or `|`

```{r}
# Your code here

```



## Optional 4 -- The Pipe in Practice

Take a numeric vector of your choice and chain at least **four** operations using `|>`. For example: create → transform → summarise → round.

Write a sentence beneath the code cell explaining what the pipe chain does in plain language.

```{r}
# Your code here

```



## Optional 5 -- Date Arithmetic

Create two `Date` objects representing politically significant events (e.g., founding of an organization, an election date, a treaty signing).

- Calculate the number of days between them
- Use `format()` to print one of the dates in a different format (e.g., `"%B %d, %Y"`)

```{r}
# Your code here

```
